### 2011-12-21 doc/theses/mikula/proposed_algorithm.tex
Master thesis: In progress.

\section{Step 1: Construction of a Syntax Tree}
The first step of the algorithm involves lexical and syntactic analyses known from compilers and produces a so-called syntax tree. The analyses are taken from Ji\v r\'{i} Schejbal's master thesis \todo{ref}. Since they are not directly related to the inference, and thus, they are not directly related to the topic of this thesis, we will not describe them. Nevertheless, they provides us with a helpful processing of XQuery queries and we can focus on the inference.

@@

\subsection{Definition of the Syntax Tree}
\begin{definition}
Syntax tree of a XQuery query $Q$ is a tuple $T = (V, E, o)$ where $V$ is a set of nodes representing XQuery language constructs (TODO\todo{kompletny zoznam}), $E$ is a set of edges - couples $(a,b):a,b \in V$ and $o:V \rightarrow \mathbb{N}$ is an ordering function and $T$ satisfies the following conditions:
\begin{itemize}
\item $(V,E)$ is a tree.
\item The root node is $ModuleNode$ representing the entire query Q.
\item $(a,b)$ is an element of $E$ if and only if $b$ is a direct component of construct $a$ in $Q$.
\item For every node $a \in V$ the ordering function $o$ is defined as follows \\* $o(a) = \left\{
	\begin{array}{ll}
		1  & \mbox{if }a\mbox{ is the root node} \\
		seq(a) & \mbox{else}
	\end{array}
\right.$ \\* where $seq(a)$ is a sequential number of appearence of construct $a$ in $Q$.
\end{itemize}
\end{definition}

\subsection{Syntax Tree Node Types}
\todo[inline]{Dat figures do appendixu?}
The node types (or classes) of the syntax are organized in an is-a hierarchical structure, commonly used in the object oriented programming languages, where a node can be of several types. This hierarchy is shown in figures \ref{FIG_syntax_tree_node_types_part_1}, \ref{FIG_syntax_tree_node_types_part_2}, \ref{FIG_syntax_tree_node_types_part_3}, \ref{FIG_syntax_tree_node_types_part_4}, \ref{FIG_syntax_tree_node_types_part_5}, \ref{FIG_syntax_tree_node_types_part_6}, \ref{FIG_syntax_tree_node_types_part_7}, \ref{FIG_syntax_tree_node_types_part_8}, \ref{FIG_syntax_tree_node_types_part_9}. Nodes with emphasised names express a common type of a group of nodes - their descendants, and these type nodes cannot be directly used in the syntax tree (abstract classes). The remaining nodes represent particular constructs of the XQuery language and their instances can be used as nodes of the syntax tree. Each of the instances additionaly contains details about the particular form of the respective XQuery construct used in the query.

>>>>

\section{Step 1: Construction of a Syntax Tree}
The first step of the algorithm involves lexical and syntax analyses known from the construction of compilers and produces a so-called syntax tree. The analyses are taken from Ji\v r\'{i} Schejbal's master thesis \todo{ref}. Since they are not directly related to the inference, and thus, they are not directly related to the topic of this thesis, we will not describe them. Nevertheless, they provides us with a helpful processing of XQuery queries and we can focus on the inference.

@@

\subsection{Definition of the Syntax Tree}
Firstly, we formally define the syntax tree.

\begin{define}[Syntax tree]
Syntax tree of XQuery query $Q$ is tuple $T = (V, E, c, \mathcal{P}, \mathcal{O})$ where $V \subset \mathbb{N}$ is a set of nodes, each node representing a particular XQuery construct in query $Q$, $E$ is a set of pairs $(v, w)$ where $v,w \in V$ and for every $a, b \in V, a \neq b: (a, b) \in E$ if and only if a construct represented by $b$ is a direct component of a construct represented by $a$ in query $Q$, $c: V \rightarrow C$ is function assigning each node with its class from set $C$ of all XQuery language constructs listed in Figures \ref{FIG_syntax_tree_node_types_part_1}, \ref{FIG_syntax_tree_node_types_part_2}, \ref{FIG_syntax_tree_node_types_part_3}, \ref{FIG_syntax_tree_node_types_part_4}, \ref{FIG_syntax_tree_node_types_part_5}, \ref{FIG_syntax_tree_node_types_part_6}, \ref{FIG_syntax_tree_node_types_part_7}, \ref{FIG_syntax_tree_node_types_part_8}, \ref{FIG_syntax_tree_node_types_part_9} (those with non-bolded names), $\mathcal{P}$ is a set of functions specifying additional properties of the nodes, and $\mathcal{O}$ is a set of functions specifying an order of children of particular nodes.
\end{define}

Regarding the additional properties, two constructs in $Q$ represented by two nodes of the same class from $C$ may differ in certain ways, and therefore, it is needed to distinguish them. For instance, two different literal values in $Q$ are represented by nodes $l_1, l_2 \in V$ and $c(l_1) = c(l_2) = LiteralNode$ but each has a different value and type. Therefore $\mathcal{P}$ contains functions \\
$type_{LiteralNode}: V_{LiteralNode} \rightarrow Types_{literal}$ \\
$value_{LiteralNode}: V_{LiteralNode} \rightarrow Values_{literal}$ \\
where $V_{LiteralNode}$ is set $\{v | v \in V, c(v) = LiteralNode\}$, $Types_{literal}$ is set of all types of literal values $\{DECIMAL, INTEGER, DOUBLE, STRING\}$, and $Values_{literal}$ is a set of all literal values (all valid XQuery decimal numbers, integers, double numbers and strings).

Set $\mathcal{P}$ contains other similar functions but due to their large count, we do not define them formally.
\todo[inline]{Moze takto ostat, alebo definovat aspon tie co pouzivame? Alebo ich aspon vypisat?}

Set $\mathcal{P}$ also contains functions that return children of particular nodes. For instance, in every syntax tree, for every $v \in V, c(v) = FLWORExprNode$ it exists $w = \in V, c(w) = WhereClauseNode$ so that $(v,w) \in E$. In other words, every $FLWORExprNode$ node has a $WhereClauseNode$ node amongst its children. Set $\mathcal{P}$ contains function \\
$whereClauseNode_{FLWORExprNode} : V_{FLWORExprNode} \rightarrow V$ \\ 
where $V_{FLWORExprNode}$ is set $\{v | v \in V, c(v) = FLWORExprNode\}$. For every $v \in V_{FLWORExprNode}: whereClauseNode_{FLWORExprNode}(v) = w$ so that $(v, w) \in E$ and $c(w) = WhereClauseNode$. Function $whereClauseNode_{FLWORExprNode}$ is a convenient function to specify a $WhereClauseNode$ child of a given $FLWORExprNode$ node.

Set $\mathcal{O}$ contains functions specifying an order of children of a some nodes. Those nodes are nodes with all children of the same class from $C$. For instance, $ParamListNode$ can has multiple $ParamNode$ children. Set $\mathcal{O}$ contains function \\
$paramOrder_{ParamListNode} : V_{ParamListNode} \rightarrow \mathcal{O}_P$ \\
where $V_{ParamListNode}$ is set $\{v | v \in V, c(v) = ParamListNode\}$ and $\mathcal{O}_P$ is set of functions $\{paramOrder_v : E_v \rightarrow \mathbb{N} | v \in V_{ParamListNode}, E_v = \{(v,w) | (v,w) \in E, c(w) = ParamNode\}\}$. For every $v \in V_{ParamListNode}: paramOrder_{ParamListNode}(v) = paramOrder_v$ and for every $e = (v, w) \in E_v: paramOrder_v(e)$ is a sequential number of $w$ amongst children of $v$ in query $Q$. In other words, it represents the order of the children as the respective constructs appear in the query.

\subsection{Syntactic Abbrevations}
Since we need to use the syntax tree in pseudo-algorithms, we define several abbrevations to make its usage more simplier.

Assuming, $v \in V$
$v.X$ is abbrevation for $X(v)$

\subsection{Closer Look at the Nodes of the Syntax Tree}
\todo[inline]{Dat figures do appendixu?}
The node types (or classes) of the syntax are organized in an is-a hierarchical structure, commonly used in the object oriented programming languages, where a node can be of several types. This hierarchy is shown in Figures \ref{FIG_syntax_tree_node_types_part_1}, \ref{FIG_syntax_tree_node_types_part_2}, \ref{FIG_syntax_tree_node_types_part_3}, \ref{FIG_syntax_tree_node_types_part_4}, \ref{FIG_syntax_tree_node_types_part_5}, \ref{FIG_syntax_tree_node_types_part_6}, \ref{FIG_syntax_tree_node_types_part_7}, \ref{FIG_syntax_tree_node_types_part_8}, \ref{FIG_syntax_tree_node_types_part_9}. Nodes with emphasised names express a common type of a group of nodes - their descendants, and these type nodes cannot be directly used in the syntax tree (abstract classes). The remaining nodes represent particular constructs of the XQuery language and their instances can be used as nodes of the syntax tree. Each of the instances additionaly contains details about the particular form of the respective XQuery construct used in the query.
